---
title: "Healthcare Data Cleaning Report"
author: "Tynisha Niles"
date: "2026-05-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(here)
library(dplyr)
library(ggplot2)
library(reshape2)
```

# 1. Dataset Overview

```{r}
data_raw <- readRDS(here("data","interim","data_ingested.rds"))
data_clean <- readRDS(here("data","processed","data_cleaned.rds"))

cat("Raw Data:\n")
cat("Rows:", nrow(data_raw), " Columns:", ncol(data_raw), "\n\n")

cat("Cleaned Data:\n")
cat("Rows:", nrow(data_clean), " Columns:", ncol(data_clean), "\n")
```

# 2. Data Quality Diagnostics (Before Cleaning)

```{r}
missing_before <- colSums(is.na(data_raw))
missing_before
```

# 3. Cleaning Decisions

The following cleaning steps were applied:

- Patient names standardised (lowercase, punctuation removed)
- Gender, Condition, Medication normalised
- Visit dates parsed into proper date format
- Phone numbers cleaned (digits only, blanks converted to NA)
- Age cleaned (text → numeric, missing values replaced with median)
- Duplicate patient records resolved using Patient Name + Visit Date

# 4. Imputation Applied

- Variable: Age  
- Method: Median imputation  
- Reason: Median reduces impact of outliers  

# 5. Post-Cleaning Summary

```{r}
summary(data_clean)
```

```{r}
missing_after <- colSums(is.na(data_clean))
missing_after
```

# 6. Charts

## 6.1 Missingness Before vs After

```{r}
missing_df <- data.frame(
  Variable = names(data_raw),
  Before = colSums(is.na(data_raw)),
  After = colSums(is.na(data_clean))
)

missing_long <- reshape2::melt(missing_df, id.vars="Variable")

ggplot(missing_long, aes(x=Variable, y=value, fill=variable)) +
  geom_bar(stat="identity", position="dodge") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title="Missingness Before vs After Cleaning")
```

## 6.2 Age Distribution

```{r}
ggplot(data_clean, aes(x=Age)) +
  geom_histogram(bins=10) +
  labs(title="Age Distribution")
```

## 6.3 Age Density

```{r}
ggplot(data_clean, aes(x=Age)) +
  geom_density() +
  labs(title="Age Density After Cleaning")
```

## 6.4 Condition Distribution

```{r}
ggplot(data_clean, aes(x=Condition)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title="Condition Frequency")
```

# 7. Limitations

- Some missing phone numbers remain  
- Median imputation may not reflect true values  
- Duplicate detection may not capture all cases  

# 8. Conclusion

The dataset has been cleaned and standardised and is ready for analysis.