---
title: "Healthcare Data Cleaning Report"
author: "Tynisha Niles"
date: "2026-05-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r libraries}
library(here)
library(dplyr)
library(ggplot2)
library(reshape2)
library(knitr)
library(tibble)
```

# 1. Dataset Overview

This report summarises the healthcare data cleaning pipeline, including diagnostics, cleaning decisions, duplicate resolution, imputation methods, and post-cleaning quality assessment.

```{r load-data}

# Raw dataset
data_raw <- readRDS(
  here(
    "data",
    "interim",
    "data_ingested.rds"
  )
)

# Final cleaned dataset
data_clean <- readRDS(
  here(
    "data",
    "processed",
    "data_cleaned.rds"
  )
)

# Row audit log
row_log <- readRDS(
  here(
    "data",
    "interim",
    "row_log.rds"
  )
)

cat("Raw Dataset:\n")
cat(
  "Rows:",
  nrow(data_raw),
  " Columns:",
  ncol(data_raw),
  "\n\n"
)

cat("Cleaned Dataset:\n")
cat(
  "Rows:",
  nrow(data_clean),
  " Columns:",
  ncol(data_clean),
  "\n"
)
```

# 2. Data Quality Diagnostics (Before Cleaning)

The raw dataset contained missing values, inconsistent formatting, categorical inconsistencies, duplicate patient visits, and mixed date formats.

```{r missing-before}

missing_before <- colSums(
  is.na(data_raw)
)

missing_before
```

# 3. Cleaning Decisions

The following cleaning procedures were applied to improve data consistency and analytical readiness:

- Patient names standardised:
  - converted to lowercase,
  - punctuation removed,
  - extra whitespace removed

- Gender values normalised into:
  - Male,
  - Female,
  - Other

- Condition and Medication fields standardised into consistent title-case formatting

- Visit dates parsed from multiple possible date formats into standard date objects

- Phone numbers cleaned by:
  - removing spaces,
  - removing non-digit characters,
  - converting blanks to missing values

- Age variable cleaned by:
  - converting text values to numeric values,
  - converting `"nan"` text values to proper missing values

- Duplicate patient visit records resolved using:
  - Patient Name
  - Visit Date

# 4. Duplicate Resolution Summary

The dataset experienced a substantial row reduction after duplicate visit resolution. Duplicate records were identified using Patient Name and Visit Date combinations.

```{r row-log}

kable(row_log)
```

The reduction in rows reflects the removal of repeated patient visit entries rather than deletion of unique patient records.

# 5. Imputation Applied

The Age variable contained missing values which were handled using median imputation.

Reason for choosing median imputation:
- robust to outliers,
- appropriate for skewed numeric variables,
- preserves central tendency.

```{r imputation-summary}

imputed_count <- sum(
  data_clean$Age_Imputed_Flag,
  na.rm = TRUE
)

median_age <- median(
  data_clean$Age,
  na.rm = TRUE
)

imputation_table <- tibble(
  Variable = "Age",
  Method = "Median Imputation",
  Values_Imputed = imputed_count,
  Median_Used = median_age
)

kable(imputation_table)
```

# 6. Post-Cleaning Summary

```{r summary-cleaned}

summary(data_clean)
```

# 7. Missingness After Cleaning

```{r missing-after}

missing_after <- colSums(
  is.na(data_clean)
)

missing_after
```

# 8. Charts and Visualisations

## 8.1 Missingness Before vs After Cleaning

```{r missingness-chart}

# Create aligned variable list

all_vars <- names(data_clean)

# Missingness BEFORE cleaning

missing_before <- sapply(
  all_vars,
  function(var) {
    
    if(var %in% names(data_raw)) {
      
      sum(is.na(data_raw[[var]]))
      
    } else {
      
      NA
    }
  }
)

# Missingness AFTER cleaning

missing_after <- sapply(
  all_vars,
  function(var) {
    
    sum(is.na(data_clean[[var]]))
  }
)

# Build comparison dataframe

missing_df <- data.frame(
  
  Variable = all_vars,
  
  Before = missing_before,
  
  After = missing_after
)

# Convert to long format

missing_long <- reshape2::melt(
  missing_df,
  id.vars = "Variable"
)

# Plot

ggplot(
  missing_long,
  aes(
    x = Variable,
    y = value,
    fill = variable
  )
) +
  geom_bar(
    stat = "identity",
    position = "dodge"
  ) +
  theme(
    axis.text.x =
      element_text(
        angle = 45,
        hjust = 1
      )
  ) +
  labs(
    title =
      "Missingness Before vs After Cleaning",
    
    x = "Variable",
    
    y = "Missing Values",
    
    fill = "Dataset"
  )

```

## 8.2 Age Distribution After Cleaning

```{r age-histogram}

ggplot(
  data_clean,
  aes(x = Age)
) +
  geom_histogram(
    bins = 10
  ) +
  labs(
    title =
      "Age Distribution After Cleaning",
    x = "Age",
    y = "Frequency"
  )
```

## 8.3 Observed vs Imputed Age Comparison

This chart compares original observed Age values against Age values after median imputation.

```{r age-comparison}

age_compare <- data.frame(
  
  Age = c(
    data_clean$Age_Original,
    data_clean$Age
  ),
  
  Type = c(
    rep(
      "Observed",
      length(data_clean$Age_Original)
    ),
    
    rep(
      "After Imputation",
      length(data_clean$Age)
    )
  )
)

ggplot(
  age_compare,
  aes(
    x = Age,
    fill = Type
  )
) +
  geom_histogram(
    alpha = 0.5,
    bins = 15,
    position = "identity"
  ) +
  labs(
    title =
      "Observed vs Imputed Age Distribution",
    x = "Age",
    y = "Frequency"
  )
```

## 8.4 Cholesterol Distribution

```{r cholesterol-chart}

ggplot(
  data_clean,
  aes(x = Cholesterol)
) +
  geom_histogram(
    bins = 10
  ) +
  labs(
    title =
      "Cholesterol Distribution",
    x = "Cholesterol",
    y = "Frequency"
  )
```

## 8.5 Condition Distribution

```{r condition-chart}

ggplot(
  data_clean,
  aes(x = Condition)
) +
  geom_bar() +
  theme(
    axis.text.x =
      element_text(
        angle = 45,
        hjust = 1
      )
  ) +
  labs(
    title =
      "Condition Frequency",
    x = "Condition",
    y = "Count"
  )
```

# 9. Limitations

Despite cleaning improvements, some limitations remain:

- Some missing phone numbers remain unresolved
- Median imputation may not reflect true patient ages
- Near-duplicate detection may not identify every duplicate case
- Some variables may still contain real-world data entry inconsistencies

# 10. Conclusion

The healthcare dataset was successfully cleaned, standardised, and prepared for downstream analysis.

Key achievements include:
- duplicate resolution,
- categorical standardisation,
- date parsing,
- missing value treatment,
- generation of audit logs,
- creation of a reproducible cleaning pipeline.

The cleaned dataset is now suitable for analysis and reporting.